A Bias-Variance Decomposition of Information Transfer in Multi-Task Learning

### 2020-08-31 arxiv/proof_main.tex
sec 7.2

	First, we provide the steps for showing the bias-variance decomposition in equation \eqref{eq_bias_multiple} and \eqref{eq_var_multiple}.
	Recall that all tasks have the same covariates in this setting.
	Therefore, the optimization objective equation \eqref{eq_mtl} becomes the following
	\begin{align}
		f(B; W_1, \dots, W_t) = \sum_{i=1}^t \bignorm{X B W_i - Y_i}^2, \label{eq_mtl_same_cov}
	\end{align}
	where $B \in \real^{p \times r}$ and $W_1, W_2, \dots, W_t \in \R^r$ for $1 < r < t$ by our assumption.
	Using the local optimality condition over $B$, that is, $\nabla_B f = 0$, we obtain a closed form solution $\hat{B}$ that depends on the output layers as follows
	\begin{align*}
		\hat{B}(W_1, \dots, W_t) &= (X^{\top}X)^{-1} X^{\top} \bigbrace{\sum_{i=1}^t Y_i W_i^{\top}} (\cal W  \cal W^{\top})^{-1} \\
		&= (B^\star \cal W ^{\top}) (\cal W \cal W ^{\top})^{-1} + (X^{\top}X)^{-1}X^{\top}   \bigbrace{\sum_{i=1}^t \varepsilon_i W_i^{\top}} (\cal W \cal W^{\top})^{-1}.
	\end{align*}
	Recall that $\hat{B}_t^{\MTL} = \hat{B} W_t$, and we obtain equation \eqref{eq_bias_multiple} and \eqref{eq_var_multiple} by a decomposition of the expected prediction loss over the random noise condition on $X$ as follows
	\begin{align*}
		 & \exarg{\varepsilon_1, \dots, \varepsilon_t}{L(\hat{B}_t^{\MTL}) \mid X} \\
		=& \exarg{\varepsilon_1, \dots, \varepsilon_t}{\bignorm{\Sigma^{1/2} \bigbrace{B^{\star} \cW^{\top} (\cW \cW^{\top})^{-1}W_t - \beta_t} + (X^{\top}X)^{-1} X^{\top} \bigbrace{\sum_{i=1}^t \varepsilon_i W_i^{\top}} (\cW \cW^{\top})^{-1} W_t}^2 \mid X} \\
		=& \bignorm{\Sigma^{1/2} (B^{\star} \cW^{\top} (\cW \cW^{\top})^{-1} W_t - \beta_t)}^2
		+ \sum_{i=1}^t \bigtr{X (X^{\top}X)^{-1}\Sigma (X^{\top}X)^{-1}X^{\top} \exarg{\varepsilon_i}{\varepsilon_i \varepsilon_i^{\top}} (W_i^{\top} (\cW \cW^{\top})^{-1} W_t)^2} \\
		=& \bignorm{\Sigma^{1/2}(B^{\star} \cW^{\top} (\cW \cW)^{-1} W_t - \beta_t)}^2
		+ \sigma^2 \cdot (W_t^{\top} (\cW \cW^{\top})^{-1} W_t) \cdot \bigtr{\Sigma (X^{\top}X)^{-1}}.
	\end{align*}
	The third step uses the fact that $\varepsilon_i$ consists of i.i.d. entries with mean zero and variance $\sigma^2$.
	The last step uses the fact that $\sum_{i=1}^t W_i W_i^{\top} = \cW \cW^{\top}$.
	One can see that the first part of the last equation is the bias and the second part is the variance.
	Hence the proof is complete.

@@

	\medskip
	\noindent\textbf{Dealing with the empirical minimizer.}
	%eet $\hat {\cal W}$ be the minimizer of $g$, and denote $\hat Q:= \hat{\cal W}^{\top} (\hat{\cal W}\hat{\cal W}^{\top})^{-1} \hat{\cal W} $.
	We show that $\norm{\hat{\cW}^{\top} (\hat{\cW}\hat{\cW}^{\top})^{-1} \cW - U_r U_r^{\top}} \le o(1)$ w.h.p.
	%\be\label{Q-Q}\|Q_0^{-1}\hat Q - \id\|_F = \oo(1) \quad \text{w.h.p.}\ee
	Since $\hat{\cW}$ is the minimizer of $g(\cdot)$, we have that
	\begin{align*}
		g(U_r) - g(\hat{\cW}) = \bigabs{g(\hat{\cW}) - g(U_r)} &= n \cdot \bigabs{\inner{{B^{\star}}^{\top} \Sigma B^{\star}}{U_r U_r^{\top} - \hat{\cW}^{\top} (\hat{\cW}\hat{\cW}^{\top})^{-1} \hat{\cW}}}\\
		&\ge n \cdot \sigma_{\min}({B^{\star}}^{\top} \Sigma B^{\star}) \cdot \bignormFro{U_r U_r^{\top} - \hat{\cW}^{\top} (\hat{\cW}\hat{\cW}^{\top})^{-1} \hat{\cW}}.
	\end{align*}
	On the other hand, by equation \eqref{eq_est_g} and triangle inequality, we have that
	\begin{align*}
		g(U_r) - g(\hat{\cW}) &\le \bigabs{g(U_r) - \ex{g(U_r)}} + (\ex{g(U_r)} - \ex{g(\hat{\cW})}) + \bigabs{g(\hat{\cW}) - \ex{g(\hat{\cW})}} \\
		&\le \bigabs{g(U_r) - \ex{g(U_r)}} +\bigabs{g(\hat{\cW}) - \ex{g(\hat{\cW})}} \\
		&\le o(n) \cdot \sigma_{\max}({B^{\star}}^{\top} \Sigma B^{\star}) + O(\sigma^2 p^{1/2 +\e}).
	\end{align*}
	Combined together, and using the assumption that the condition number of ${B^{\star}}^{\top}\Sigma B^{\star}$ does not grow with $p$ and $\sigma_{\min}({B^{\star}}^{\top} \Sigma B^{\star}) \ge \sigma^2 / p^{1/2}$, we conclude that the distance between the two subspaces $\hat{\cW}^{\top}(\hat{\cW}\hat{\cW}^{\top})^{-1}\hat{\cW}$ and $U_r U_r^{\top}$ diminish as $p$ grows.
	%The second step uses the fact that $U_r U_r^{\top}$ is the global minimum of $\ex{g(\cdot)}$.
	%In fact, if \eqref{Q-Q} does not hold, then using the condition $\lambda_{\min}((B^{\star})^\top\Sigma B^{\star})\gtrsim \sigma^2$ and that $\delta_{\vari}(\cal W)=\OO(\sigma^2)$ by  Lemma \ref{lem_minv}, we obtain that
	%$$   \val(\hat {Q}) + t  \sigma^2 > (\val( Q_0) + t \sigma^2 )\cdot (1+\oo(1)) \ \Rightarrow \ g( \hat{\cal W})>g( {\cal W}_0),$$
	%where $\cal W_0\in \R^{r\times t}$ is a matrix such that  $ \cal W_0^{\top} (\cal W_0\cal W_0^{\top})^{-1} \cal W_0=Q_0$. Hence $\hat {\cal W}$ is not a minimizer, which leads to a contradiction.

	\medskip
	\noindent\textbf{Putting both steps together.}
	In sum, we have solved that $\hat{\beta}_i^{\MTL}=B^{\star}\left( U_r v_i +\oo(1)\right)$. Inserting it into the definition of the test loss, we get that with high probability,
	\begin{align*}
		\te(\hat{\beta}_t^{\MTL}) &= \bignorm{\Sigma^{1/2} \left((B^\star \hat{\cal W}^\top) (\hat{\cal W}\hat{\cal W}^{\top})^{-1} \hat W_t - \beta_t \right) }^2
		+ \sigma^2  \hat W_t^{\top} (\hat{\cal W}\hat{\cal W}^{\top})^{-1} \hat W_t \cdot \bigtr{\Sigma (X^{\top}X)^{-1}} \\
		&= \bignorm{\Sigma^{1/2} \bigbrace{B^{\star} U_r v_t-\beta_t}}^2 + \oo\left(\|B^\star\|^2\right) + \sigma^2\norm{v_t}^2 \bigtr{\Sigma (X^{\top}X)^{-1}} \cdot (1+\oo(1))\\
		&= \bignorm{\Sigma^{1/2} \bigbrace{B^{\star} U_r v_t-\beta_t}}^2 + \frac{\sigma^2}{\rho-1}\norm{v_t}^2 + \oo \left( \|B^\star\|^2 + \sigma^2\right),
	\end{align*}
 where we used Lemma \ref{lem_minv} in the last step. On the other hand, by Lemma \ref{lem_minv} we have
	$$\te(\hat{\beta}_t^{\STL})=\frac{\sigma^2}{\rho-1} \cdot \left( 1+\oo(1)\right).$$
Combining the above two estimates, we conclude the proof.
\end{proof}
%From the above we can obtain three conceptual insights that are consistent with Section \ref{sec_denoise} and \ref{sec_insight}.
%\begin{itemize}
%	\item The de-noising effect of multi-task learning.
%	\item Multi-task training vs single-task training can be either positive or negative.
%	\item Transfer learning is better than the other two. And the improvement over multi-task training increases as the model distances become larger.
%\end{itemize}

>>>>

	First, we provide the steps for showing the bias-variance decomposition in equation \eqref{eq_bias_multiple} and \eqref{eq_var_multiple}.
	Recall that all tasks have the same covariates in this setting.
	Therefore, the optimization objective equation \eqref{eq_mtl} becomes the following
	\begin{align}
		f(B; W_1, \dots, W_t) = \sum_{i=1}^t \bignorm{X B W_i - Y_i}^2, \label{eq_mtl_same_cov}
	\end{align}
	where $B \in \real^{p \times r}$ and $W_1, W_2, \dots, W_t \in \R^r$ for $1 < r < t$ by our assumption.
	Using the local optimality condition over $B$, that is, $\nabla_B f = 0$, we obtain a closed form solution $\hat{B}$ that depends on the output layers as follows
	\begin{align*}
		\hat{B}(W_1, \dots, W_t) &= (X^{\top}X)^{-1} X^{\top} \bigbrace{\sum_{i=1}^t Y_i W_i^{\top}} (\cal W  \cal W^{\top})^{-1} \\
		&= (B^\star \cal W ^{\top}) (\cal W \cal W ^{\top})^{-1} + (X^{\top}X)^{-1}X^{\top}   \bigbrace{\sum_{i=1}^t \varepsilon_i W_i^{\top}} (\cal W \cal W^{\top})^{-1}.
	\end{align*}
	Recall that $\hat{B}_t^{\MTL} = \hat{B} W_t$, and we obtain equation \eqref{eq_bias_multiple} and \eqref{eq_var_multiple} by a decomposition of the expected prediction loss over the random noise condition on $X$ as follows
	\begin{align}
		 & \exarg{\varepsilon_1, \dots, \varepsilon_t}{L(\hat{B}_t^{\MTL}) \mid X} \nonumber \\
		=& \exarg{\varepsilon_1, \dots, \varepsilon_t}{\bignorm{\Sigma^{1/2} \bigbrace{B^{\star} \cW^{\top} (\cW \cW^{\top})^{-1}W_t - \beta_t} + (X^{\top}X)^{-1} X^{\top} \bigbrace{\sum_{i=1}^t \varepsilon_i W_i^{\top}} (\cW \cW^{\top})^{-1} W_t}^2 \mid X} \nonumber \\
		=& \bignorm{\Sigma^{1/2} (B^{\star} \cW^{\top} (\cW \cW^{\top})^{-1} W_t - \beta_t)}^2
		+ \sum_{i=1}^t \bigtr{X (X^{\top}X)^{-1}\Sigma (X^{\top}X)^{-1}X^{\top} \exarg{\varepsilon_i}{\varepsilon_i \varepsilon_i^{\top}} (W_i^{\top} (\cW \cW^{\top})^{-1} W_t)^2} \nonumber \\
		=& \bignorm{\Sigma^{1/2}(B^{\star} \cW^{\top} (\cW \cW)^{-1} W_t - \beta_t)}^2
		+ \sigma^2 \cdot (W_t^{\top} (\cW \cW^{\top})^{-1} W_t) \cdot \bigtr{\Sigma (X^{\top}X)^{-1}}. \label{eq_ex_pred}
	\end{align}
	The third step uses the fact that $\varepsilon_i$ consists of i.i.d. entries with mean zero and variance $\sigma^2$.
	The last step uses the fact that $\sum_{i=1}^t W_i W_i^{\top} = \cW \cW^{\top}$.
	One can see that the first part of the last equation is the bias and the second part is the variance.
	Hence the proof is complete.

@@

	\medskip
	\noindent\textbf{Dealing with the empirical minimizer.}
	%eet $\hat {\cal W}$ be the minimizer of $g$, and denote $\hat Q:= \hat{\cal W}^{\top} (\hat{\cal W}\hat{\cal W}^{\top})^{-1} \hat{\cal W} $.
	We show that $\norm{\hat{\cW}^{\top} (\hat{\cW}\hat{\cW}^{\top})^{-1} \cW - U_r U_r^{\top}} \le o(1)$ w.h.p.
	%\be\label{Q-Q}\|Q_0^{-1}\hat Q - \id\|_F = \oo(1) \quad \text{w.h.p.}\ee
	Since $\hat{\cW}$ is the minimizer of $g(\cdot)$, we have that
	\begin{align*}
		g(U_r) - g(\hat{\cW}) = \bigabs{g(\hat{\cW}) - g(U_r)} &= n \cdot \bigabs{\inner{{B^{\star}}^{\top} \Sigma B^{\star}}{U_r U_r^{\top} - \hat{\cW}^{\top} (\hat{\cW}\hat{\cW}^{\top})^{-1} \hat{\cW}}}\\
		&\ge n \cdot \sigma_{\min}({B^{\star}}^{\top} \Sigma B^{\star}) \cdot \bignormFro{U_r U_r^{\top} - \hat{\cW}^{\top} (\hat{\cW}\hat{\cW}^{\top})^{-1} \hat{\cW}}.
	\end{align*}
	On the other hand, by equation \eqref{eq_est_g} and triangle inequality, we have that
	\begin{align*}
		g(U_r) - g(\hat{\cW}) &\le \bigabs{g(U_r) - \ex{g(U_r)}} + (\ex{g(U_r)} - \ex{g(\hat{\cW})}) + \bigabs{g(\hat{\cW}) - \ex{g(\hat{\cW})}} \\
		&\le \bigabs{g(U_r) - \ex{g(U_r)}} +\bigabs{g(\hat{\cW}) - \ex{g(\hat{\cW})}} \\
		&\le o(n) \cdot \sigma_{\max}({B^{\star}}^{\top} \Sigma B^{\star}) + O(\sigma^2 p^{1/2 +\e}).
	\end{align*}
	Combined together, and using the assumption that the condition number of ${B^{\star}}^{\top}\Sigma B^{\star}$ does not grow with $p$ and $\sigma_{\min}({B^{\star}}^{\top} \Sigma B^{\star}) \ge \sigma^2 / p^{1/2}$, we conclude that the distance between the two subspaces $\hat{\cW}^{\top}(\hat{\cW}\hat{\cW}^{\top})^{-1}\hat{\cW}$ and $U_r U_r^{\top}$ diminish to $o(1)$ as $p$ grows.
	%The second step uses the fact that $U_r U_r^{\top}$ is the global minimum of $\ex{g(\cdot)}$.
	%In fact, if \eqref{Q-Q} does not hold, then using the condition $\lambda_{\min}((B^{\star})^\top\Sigma B^{\star})\gtrsim \sigma^2$ and that $\delta_{\vari}(\cal W)=\OO(\sigma^2)$ by  Lemma \ref{lem_minv}, we obtain that
	%$$   \val(\hat {Q}) + t  \sigma^2 > (\val( Q_0) + t \sigma^2 )\cdot (1+\oo(1)) \ \Rightarrow \ g( \hat{\cal W})>g( {\cal W}_0),$$
	%where $\cal W_0\in \R^{r\times t}$ is a matrix such that  $ \cal W_0^{\top} (\cal W_0\cal W_0^{\top})^{-1} \cal W_0=Q_0$. Hence $\hat {\cal W}$ is not a minimizer, which leads to a contradiction.

	\medskip
	\noindent\textbf{Putting all parts together.}
	%In sum, we have solved that $\hat{\beta}_i^{\MTL}=B^{\star}\left( U_r v_i +\oo(1)\right)$.
	Now we are ready to finish the proof.
	Similar to the proof of equation \eqref{eq_est_g1} and \eqref{eq_est_g2}, we have that
	\begin{align}
		L(\hat{\beta}_t^{\MTL}) = \exarg{\varepsilon_1, \dots, \varepsilon_t}{L(\hat{\beta}_t^{\MTL})} (1 + O(p^{-1/2+c})) + O(\sigma^2 \cdot p^{-1/2 + c}). \label{eq_multiple_last1}
	\end{align}
	Using equation \eqref{eq_ex_pred}, we have that
	\begin{align}
		\exarg{\varepsilon_1, \dots, \varepsilon_t}{\hat{\beta}_t^{\MTL}} &= \bignorm{\Sigma^{1/2} \left((B^\star \hat{\cal W}^\top) (\hat{\cal W}\hat{\cal W}^{\top})^{-1} \hat W_t - \beta_t \right) }^2
		+ \sigma^2  \hat W_t^{\top} (\hat{\cal W}\hat{\cal W}^{\top})^{-1} \hat W_t \cdot \bigtr{\Sigma (X^{\top}X)^{-1}} \nonumber \\
		&= \bignorm{\Sigma^{1/2} \bigbrace{B^{\star} U_r v_t-\beta_t}}^2 + \oo\left(\bigtr{{B^{\star}}^{\top} \Sigma B^{\star}}\right) + \sigma^2(\norm{v_t}^2 + o(1)) \bigtr{\Sigma (X^{\top}X)^{-1}} \cdot (1+\oo(1)) \nonumber \\
		&= \bignorm{\Sigma^{1/2} \bigbrace{B^{\star} U_r v_t-\beta_t}}^2 + \frac{\sigma^2}{\rho-1}\norm{v_t}^2 + \oo \left( \bigtr{{B^{\star}}^{\top} \Sigma B^{\star}} + \sigma^2\right), \label{eq_multiple_last2}
	\end{align}
	In the second step, we use that $\norm{\hat{\cW}^{\top} (\cW \cW^{\top})^{-1} \cW - U_r U_r^{\top}} \le o(1)$ w.h.p., which implies that
	\[ \bignorm{U_r v_t - \hat{\cW}^{\top} (\cW \cW^{\top})^{-1} \hat{\cW}_t} \le o(1) \quad \text{ and } \quad \bigabs{\hat{\cW}_t^{\top} (\hat{\cW} \hat{\cW}^{\top})^{-1} \hat{W}_t - \norm{v_t}^2} \le o(1). \]
  In the last step, we use Lemma \ref{lem_minv}, that is, $\bigtr{\Sigma (X^{\top} X)^{-1}} = \sigma^2 \cdot (1 + o(1))/ (\rho - 1)$.
	Combining equation \eqref{eq_multiple_last1}, \eqref{eq_multiple_last2}, and $\te(\hat{\beta}_t^{\STL})=\frac{\sigma^2}{\rho-1} \cdot \left( 1+\oo(1)\right)$, we conclude the proof.
\end{proof}
%From the above we can obtain three conceptual insights that are consistent with Section \ref{sec_denoise} and \ref{sec_insight}.
%\begin{itemize}
%	\item The de-noising effect of multi-task learning.
%	\item Multi-task training vs single-task training can be either positive or negative.
%	\item Transfer learning is better than the other two. And the improvement over multi-task training increases as the model distances become larger.
%\end{itemize}
